A Parallel General Purpose Multi-Objective Optimization Framework, with Application To Electron Beam Dynamics

### 2018-09-29 paper/prab.tex
remove cite
add linenumbers
taking out we's

\begin{abstract}
Particle accelerators are invaluable tools for research in the basic and
  applied sciences, in fields such as materials science, chemistry,
  the biosciences, particle physics, nuclear physics and medicine.
The design, commissioning, and operation of accelerator facilities is a
  non-trivial task, due to the large number of control parameters and the
  complex interplay of several conflicting design goals.
We propose to tackle this problem by means of multi-objective optimization
  algorithms which also facilitate a parallel deployment.
In order to compute solutions in a meaningful time frame we require a fast
  and scalable software framework.
In this paper, we present the implementation of such a general-purpose
  framework for simulation-based multi-objective optimization methods that
  allows the automatic investigation of optimal sets of machine parameters.
The implementation is based on a master/slave paradigm, employing several
  masters that govern a set of slaves executing simulations and performing
  optimization tasks.
Using evolutionary algorithms and \textsc{OPAL}~simulations as optimizer and
  forward solver in our framework, we present validation experiments and first
  results of multi-objective optimization problems in the domain of beam
  dynamics.

@@

A modular multi-objective software framework was developed (see
 Fig.~\ref{fig:framenetwork}) where the core functionality of the optimizer is decoupled from
 the ``beam dynamics'' but fully integrated in the OPAL \cite{opal} framework. 
To that end, we use a master/slave mechanism where a master process governs a
 set of slave processes given some computational tasks (beam dynamics simulation) to complete.
This separation allows easy interchange of optimization algorithms, forward
  solvers and optimization problems.
A ``pilot'' coordinates all efforts between the optimization algorithm and the
  beam dynamics task. In the sequel we will also use the notion of ``forward solver'' to indicate 
  the beam dynamics task.
This forms a robust and general framework for massively parallel
  multi-objective optimization.
Currently the framework offers one concrete optimization algorithm, an
  evolutionary algorithm employing a \textsc{NSGA-II} selector \cite{pisa}.
Normally, simulation based approaches are plagued by the trade-of between
  level of detail and time to solution.
We address this problem later in Section~(\ref{awa:subsection:test}) by using forward solvers with different time and
  resolution complexity.

@@

Once the shape of the Pareto front has been determined, the buyer can
  \textit{specify preference}, balancing the features by observing the
  effect on the optimality criteria, converging to the preferred solution.
This is called \textit{a posteriori} preference specification since we select
  a solution after all possible trade-offs have been presented to us.
An alternative is to specify preference \textit{a priori}, e.g. by
  weighting (specifying preference before solving the problem) and combining
  all objectives into one and applying a single-objective method to solve the
  problem (yielding only one solution).
In many situations preference is not known \textit{a priori} and an
  \textit{a posteriori} preference specification helps conveying a deeper
  understanding of the solution space.
The Pareto front can be explored and the impact of a trade-off decision then
  becomes visible.

Sampling Pareto fronts is far from trivial.
A number of approaches have been proposed,
  e.g.\ evolutionary algorithms~\cite{deb:09},
  simulated annealing~\cite{kigv:83},
  swarm methods~\cite{keeb:95},
  and many more~\cite{domc:96,cati:02,kara:05,hoss:09}.
In the next section we briefly introduce the theory of evolutionary algorithms
  used in the present work.

@@

Since there already exist plenty of implementations of evolutionary algorithms,
  we decided to incorporate the PISA library \cite{pisa} into our
  framework.
One of the advantages of PISA is that it separates variator from selector,
  rendering the library expandable and configurable.
Implementing a variator was enough to use PISA in our framework and
  retain access to all available PISA selectors.
As shown in Fig.~\ref{fig:varsel}, the selector is in charge of ordering a
  set of $d$-dimensional vectors and selecting the $k$ fittest individuals
  currently in the population.
The performance of a selector depends on the number of objectives and the
  surface of the search space.
So far, the NSGA-II selector \cite{dpam:02} has been used and exhibits satisfactory
  convergence performance.

The task of the variator is to generate offspring and ensure diversity in the
  population.
The variator can start generating offspring once the fitness of every
  individual of the population has been evaluated.
This explicit synchronization point defines an obvious bottleneck for parallel
  implementations of evolutionary algorithms.
In the worst case some MPI processes are taking a long time to compute the
  fitness of the last individual in the pool of individuals to evaluate.
During this time all other resources are idle and wait for the result of
  this one individual in order to continue to generate and evaluate offspring.
To counteract this effect we call the selector already when two individuals
  have finished evaluating their fitness, lifting the boundaries between
  generations and evaluating the performance of individuals.
New offspring will be generated and MPI processes can immediately return to
  work on the next fitness evaluation.
By calling the selector more frequently (already after two offspring
  individuals have been evaluated) results in better populations since bad
  solutions are rejected earlier.
On the other hand, calling the selector more often is computationally more
  expensive.

@@

Several similar frameworks, e.g.~\cite{fide:09,lems:09,lbjt:07,dnld:06}, have
  been proposed.
Commonly these frameworks are tightly coupled to an optimization algorithm,
  e.g.\ only providing evolutionary algorithms as optimizers.
Users can merely specify optimization problems, but cannot change the
  optimization algorithm.
Our framework follows a more general approach, providing a user-friendly way
  to introduce new or choose from existing built-in multi-objective
  optimization algorithms.
Tailoring the optimization algorithm to the optimization problem at hand is
  an important feature due to the many different characteristics of
  optimization problems that should be handled by such a general framework.
As an example, we show how \textsc{Pisa} \cite{pisa}, an existing evolutionary
  algorithm library, was integrated with ease.
Similarly, other multi-objective algorithms could be incorporated and
  used to solve optimization problems.

@@

The basic assumption in simulation-based optimization is that we need to
  call an expensive simulation software component present in the
  constraints or objectives.
We divide the framework in three exchangeable components, as shown in
  Fig.~\ref{fig:opt-framework-layout}, to encapsulate the major behavioral
  patterns of the framework.
%
\begin{figure}
  \centering
  \scalebox{0.85}{
  \begin{tikzpicture}[text=black]
  \input{figures/opt-framework-layout.tex}
  \end{tikzpicture}
}

@@

In most cases forward solver implementations are simple wrappers to run
  an existing ``external'' simulation code using a set of design variables as
  input. In case of the OPAL integration, basically the \texttt{main} function is
  playing the role of the ``forward solver''. Underlying the general nature of our approach, 
  in a similar project, we use the described methods for cavity shape optimisation based on \cite{ARBENZ2008381}. 
As for the \textsc{Optimizer} component there exists a base class, labeled
  \texttt{Simulation} as common basis for all \textsc{Simulation}
  implementations.
In addition, this component also inherits from the \texttt{Worker} class,
  already implementing the polling protocol for default worker types.
As shown in the API in Listing 3, the \texttt{Worker} class expects an
  implementation to provide implementations for those three methods.

@@

To allow arbitrary constraints and objective expressions, such as
%
%{0.2cm}
\begin{Verbatim}[fontsize=\scriptsize]
  name: OBJECTIVE,
        EXPR="5 * average(42.0, "measurement.dat") + ENERGY";
\end{Verbatim}
%\vspace{0.2cm}
%
\noindent
  we implemented an expression parser using Boost Spirit~\cite{boost}.
In addition to the parser, we need an evaluator able to evaluate an expression,
  given a parse tree and variable assignments to an actual value.
Expressions arising in multi-objective optimization problems usually evaluate
  to booleans or floating point values.
The parse tree, also denoted abstract syntax tree (AST), is constructed
  recursively while an expression is parsed.
Upon evaluation, all unknown variables are replaced with values, either
  obtained from simulation results or provided by other subtrees in the AST.
In this stage, the AST can be evaluated bottom-up and the desired result is
  returned after processing the root of the tree.

To improve the expressive power of objectives and constraints we introduced a
  simple mechanism to define and call custom functions in expressions.
Using simple functors, e.g. as the one shown in Listing 4 to compute an
  average over a set of data points, enriches expressions with custom
  functions.
Custom function implementations overload the \texttt{()} parenthesis operator.
The function arguments specified in the corresponding expression are stored in
  a \texttt{std::vector} of Boost variants~\cite{boost2} that can be
  booleans, strings or floating point values.

@@

A general-purpose framework for solving multi-objective
  optimization problems was presented.
Its modular design simplifies the application to simulation-based optimization
  problems for a wide range of problems and allows to exchange the
  optimization algorithm.
In a recent work, we successfully implemented and integrated another
  optimization algorithm, originally presented in~\cite{pesc:11}, into our
  framework.
The flexibility of being able to adapt both ends of the optimization
  process, the forward solver and the optimization algorithm simultaneously
  not only leads to broad applicability but it facilitates the ability to
  tailor the optimization strategy to the optimization problem as well.

>>>>

\begin{abstract}
Particle accelerators are invaluable tools for research in the basic and
  applied sciences, in fields such as materials science, chemistry,
  the biosciences, particle physics, nuclear physics and medicine.
The design, commissioning, and operation of accelerator facilities is a
  non-trivial task, due to the large number of control parameters and the
  complex interplay of several conflicting design goals.
We propose to tackle this problem by means of multi-objective optimization
  algorithms which also facilitate a parallel deployment.
In order to compute solutions in a meaningful time frame a fast 
  and scalable software framework is required.
In this paper, we present the implementation of such a general-purpose
  framework for simulation-based multi-objective optimization methods that
  allows the automatic investigation of optimal sets of machine parameters.
The implementation is based on a master/slave paradigm, employing several
  masters that govern a set of slaves executing simulations and performing
  optimization tasks.
Using evolutionary algorithms and \textsc{OPAL}~simulations as optimizer and
  forward solver in our framework, validation experiments and first
  results of multi-objective optimization problems in the domain of beam
  dynamics is presented.

@@

A modular multi-objective software framework was developed (see
 Fig.~\ref{fig:framenetwork}) where the core functionality of the optimizer is decoupled from
 the ``beam dynamics'' but fully integrated in the OPAL \cite{opal} framework. 
To that end, we use a master/slave mechanism where a master process governs a
 set of slave processes given some computational tasks (beam dynamics simulation) to complete.
This separation allows easy interchange of optimization algorithms, forward
  solvers and optimization problems.
A ``pilot'' coordinates all efforts between the optimization algorithm and the
  beam dynamics task. In the sequel we will also use the notion of ``forward solver'' to indicate the beam dynamics task.
This forms a robust and general framework for massively parallel
  multi-objective optimization.
Currently the framework offers one concrete optimization algorithm, an
  evolutionary algorithm employing a \textsc{NSGA-II} selector \cite{pisa}.
Normally, simulation based approaches are plagued by the trade-of between
  level of detail and time to solution.
This problem is addressed later in Section~(\ref{awa:subsection:test}) by using forward solvers with different time and
  resolution complexity.

@@

Once the shape of the Pareto front has been determined, the buyer can
  \textit{specify preference}, balancing the features by observing the
  effect on the optimality criteria, converging to the preferred solution.
This is called \textit{a posteriori} preference specification since 
  a solution is selected after all possible trade-offs have been presented to us.
An alternative is to specify preference \textit{a priori}, e.g. by
  weighting (specifying preference before solving the problem) and combining
  all objectives into one and applying a single-objective method to solve the
  problem (yielding only one solution).
In many situations preference is not known \textit{a priori} and an
  \textit{a posteriori} preference specification helps conveying a deeper
  understanding of the solution space.
The Pareto front can be explored and the impact of a trade-off decision then
  becomes visible.

Sampling Pareto fronts is far from trivial.
A number of approaches have been proposed,
  e.g.\ evolutionary algorithms~\cite{deb:09},
  simulated annealing~\cite{kigv:83},
  swarm methods~\cite{keeb:95},
  and many more~\cite{domc:96,cati:02,kara:05,hoss:09}.
In the next section, we briefly introduce the theory of evolutionary algorithms
  used in the present work.

@@

Since there already exist plenty of implementations of evolutionary algorithms,
  it was decided to incorporate the PISA library \cite{pisa} into our
  framework.
One of the advantages of PISA is that it separates variator from selector,
  rendering the library expandable and configurable.
Implementing a variator was enough to use PISA in our framework and
  retain access to all available PISA selectors.
As shown in Fig.~\ref{fig:varsel}, the selector is in charge of ordering a
  set of $d$-dimensional vectors and selecting the $k$ fittest individuals
  currently in the population.
The performance of a selector depends on the number of objectives and the
  surface of the search space.
So far, the NSGA-II selector \cite{dpam:02} has been used and exhibits satisfactory
  convergence performance.

The task of the variator is to generate offspring and ensure diversity in the
  population.
The variator can start generating offspring once the fitness of every
  individual of the population has been evaluated.
This explicit synchronization point defines an obvious bottleneck for parallel
  implementations of evolutionary algorithms.
In the worst case some MPI processes are taking a long time to compute the
  fitness of the last individual in the pool of individuals to evaluate.
During this time all other resources are idle and wait for the result of
  this one individual in order to continue to generate and evaluate offspring.
To counteract this effect, the selector is already called when two individuals
  have finished evaluating their fitness, lifting the boundaries between
  generations and evaluating the performance of individuals.
New offspring will be generated and MPI processes can immediately return to
  work on the next fitness evaluation.
By calling the selector more frequently (already after two offspring
  individuals have been evaluated) results in better populations since bad
  solutions are rejected earlier.
On the other hand, calling the selector more often is computationally more
  expensive.

@@

Several similar frameworks, e.g.~\cite{fide:09,lems:09,lbjt:07,dnld:06}, have
  been proposed.
Commonly these frameworks are tightly coupled to an optimization algorithm,
  e.g.\ only providing evolutionary algorithms as optimizers.
Users can merely specify optimization problems, but cannot change the
  optimization algorithm.
Our framework follows a more general approach, providing a user-friendly way
  to introduce new or choose from existing built-in multi-objective
  optimization algorithms.
Tailoring the optimization algorithm to the optimization problem at hand is
  an important feature due to the many different characteristics of
  optimization problems that should be handled by such a general framework.
As an example, it is shown how \textsc{Pisa} \cite{pisa}, an existing evolutionary
  algorithm library, was integrated with ease.
Similarly, other multi-objective algorithms could be incorporated and
  used to solve optimization problems.

@@

The basic assumption in simulation-based optimization is that a
  call to an expensive simulation software component present in the
  constraints or objectives is needed.
The framework is divided in three exchangeable components, as shown in
  Fig.~\ref{fig:opt-framework-layout}, to encapsulate the major behavioral
  patterns of the framework.
%
\begin{figure}
  \centering
  \scalebox{0.85}{
  \begin{tikzpicture}[text=black]
  \input{figures/opt-framework-layout.tex}
  \end{tikzpicture}
}

@@

In most cases forward solver implementations are simple wrappers to run
  an existing ``external'' simulation code using a set of design variables as
  input. In case of the OPAL integration, basically the \texttt{main} function is
  playing the role of the ``forward solver''. Underlying the general nature of our approach, 
  in a similar project, the described methods are used for cavity shape optimisation based on \cite{ARBENZ2008381}. 
As for the \textsc{Optimizer} component there exists a base class, labeled
  \texttt{Simulation} as common basis for all \textsc{Simulation}
  implementations.
In addition, this component also inherits from the \texttt{Worker} class,
  already implementing the polling protocol for default worker types.
As shown in the API in Listing 3, the \texttt{Worker} class expects an
  implementation to provide implementations for those three methods.

@@

To allow arbitrary constraints and objective expressions, such as
%
%{0.2cm}
\begin{Verbatim}[fontsize=\scriptsize]
  name: OBJECTIVE,
        EXPR="5 * average(42.0, "measurement.dat") + ENERGY";
\end{Verbatim}
%\vspace{0.2cm}
%
\noindent
  An expression parser using Boost Spirit~\cite{boost} was implemented.
In addition to the parser, we need an evaluator able to evaluate an expression,
  given a parse tree and variable assignments to an actual value.
Expressions arising in multi-objective optimization problems usually evaluate
  to booleans or floating point values.
The parse tree, also denoted abstract syntax tree (AST), is constructed
  recursively while an expression is parsed.
Upon evaluation, all unknown variables are replaced with values, either
  obtained from simulation results or provided by other subtrees in the AST.
In this stage, the AST can be evaluated bottom-up and the desired result is
  returned after processing the root of the tree.

To improve the expressive power of objectives and constraints, a
  simple mechanism to define and call custom functions in expressions was introduced.
Using simple functors, e.g. as the one shown in Listing 4 to compute an
  average over a set of data points, enriches expressions with custom
  functions.
Custom function implementations overload the \texttt{()} parenthesis operator.
The function arguments specified in the corresponding expression are stored in
  a \texttt{std::vector} of Boost variants~\cite{boost2} that can be
  booleans, strings or floating point values.

@@

A general-purpose framework for solving multi-objective
  optimization problems was presented.
Its modular design simplifies the application to simulation-based optimization
  problems for a wide range of problems and allows to exchange the
  optimization algorithm.
%In a recent work, we successfully implemented and integrated another
  %optimization algorithm, originally presented in~\cite{pesc:11}, into our
 % framework.
The flexibility of being able to adapt both ends of the optimization
  process, the forward solver and the optimization algorithm simultaneously
  not only leads to broad applicability but it facilitates the ability to
  tailor the optimization strategy to the optimization problem as well.
